---
title: "Lesson 10: Wrangling Character Strings with stringr"
author: "Gabriel Odom, Catalina Cañizares, and Posit Contributors"
date: "07/07/2022"
date-modified: "2023-11-28"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # , eval = FALSE
```

# Review
What did we learn last class? 

1. Subset Rows with `filter()`
2. Sort Rows with `arrange()`
3. Select and Move Columns with `select()`
4. Create New Columns with `mutate()`
5. Grouping and Group Summaries with `group_by()` and `summarise()`
6. Using `*_join()` to Merge Tibbles

</br>

*******************************************************************************
</br>



# Overview
In this lesson, we will cover the following:

1. String Matching
2. Modifying Strings
3. Substring Extraction
4. Trim and Pad Strings


## More About this Lesson
The original version of this material was largely from my memory and what Catalina and I needed to solve some problems, but "version 2" was restructured to draw from the training materials here: <https://rstudio.github.io/cheatsheets/html/strings.html>. The cool thing about the `stringr` package is that all of the functions start with `str_`. This means that you can more easily find helpful string functions. Also, as with all of the packages in the `tidyverse`, the `stringr` package comes with a nice cheat sheet: <https://rstudio.github.io/cheatsheets/strings.pdf>.


## Example Data
We will use two data sets as examples in this lesson, one easy and one complex.

- Easy: the `fruit` object from the `stringr` package. This is a simple character vector of names of different fruits. This small data set comes automatically with the Tidyverse.
- Complex: the `outcomesCTN0094` data frame, with column `usePatternUDS`, from the `CTNote` package. Install this package via (make sure to uncomment the install line the first time you run it)
```{r packages, message=FALSE}
# install.packages("CTNote")
library(CTNote)
library(tidyverse)

data("outcomesCTN0094")
```

::: {.callout-note title="Exercises"}
## Exercises
1. Inspect the vector `fruit`.
2. Create a smaller version of the `outcomesCTN0094` tibble with the following columns: `who`, `usePatternUDS`, and `ctn0094_relapse_time`. Save it as an object in your Global environment called `outcome_df`.
:::

```{r answer-to-exercise}
#| echo: false

outcome_df <- 
  outcomesCTN0094 %>% 
  select(who, usePatternUDS, ctn0094_relapse_time)
rm(outcomesCTN0094)
```

</br>

*******************************************************************************
</br>



# String Matching


## Functions to Know

- `str_detect()`: detect the presence/absence of a matching pattern
- `str_count()`: count number of times a matching pattern appears in a string


## Use Cases
In the `fruit` vector, we may want to find which fruit names have the word "berry" or "berries" in them, then print those names. Because I want to detect both, I have two options.

Option 1: the character intersection of "berry" and "berries":
```{r find-berries-1}
# Create a logical vector to indicate which strings have the matching pattern
str_detect(string = fruit, pattern = "berr")

# Print the names of the fruits which have the matching pattern
fruit[ str_detect(string = fruit, pattern = "berr") ]
```

Option 2: using an "OR" statement (the `|` symbol):
```{r find-berries-2}
# Create a logical vector to indicate which strings have the matching pattern
str_detect(string = fruit, pattern = "berry|berries")

# Print the names of the fruits which have the matching pattern
fruit[ str_detect(string = fruit, pattern = "berry|berries") ]
```


</br>

*******************************************************************************
</br>



# Modifying Strings


## Functions to Know

- `str_replace()` / `str_replace_all()`: replace the first/all matches of a pattern in a string with new text
- `str_remove()` / `str_remove_all`: remove the first/all matches of a pattern in a string
- `str_to_lower()` / `str_to_upper()`: convert a string to lower case/UPPER CASE

</br>

*******************************************************************************
</br>



# Substring Extraction


## Functions to Know

- `str_sub()`: extracts or replaces the elements at **a single position** in each string
- `str_extract()` / `str_extract_all()`: extract the first/all matches of a pattern from each string

</br>

*******************************************************************************
</br>



# Trim and Pad Strings


## Functions to Know

- `str_length()`: compute the number of "letters" in a string
- `str_trim()`: remove spaces from the start and end of string
- `str_pad()`: add spaces (or some other "letter") to the start or end of a string until the string is a specified length

</br>

*******************************************************************************
</br>



# Modifying Strings in Tibbles
So far, we have seen how the `stringr` functions can be used on a single character vector. However, many times we have character information stored as a column of a tibble. This requires us to modidy our approach.
