---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(palmerpenguins)
library(tidyverse)

data("penguins")
```


## Introduction to lab

We are going to work with the data set *penguins_raw* from the
`palmerpinguins` package. 

Data set description: 
*A tibble with 344 rows and 17 variables*


| Variable Name     | Description                                                                             |
| ----------------- | --------------------------------------------------------------------------------------- |
| studyName         | Sampling expedition from which data were collected, generated, etc.                     |
| Sample Number     | an integer denoting the continuous numbering sequence for each sample                   |
| Species           | a character string denoting the penguin species                                         |
| Region            | a character string denoting the region of Palmer LTER sampling grid                     |
| Island            | a character string denoting the island near Palmer Station where samples were collected |
| Stage             | a character string denoting reproductive stage at sampling                              |
| Individual ID     | a character string denoting the unique ID for each individual in dataset                |
| Clutch Completion | a character string denoting if the study nest observed with a full clutch, i.e., 2 eggs |
| Date Egg          | a date denoting the date study nest observed with 1 egg (sampled)                       |
| Culmen Length     | a number denoting the length of the dorsal ridge of a bird's bill (millimeters)         |
| Culmen Depth      | a number denoting the depth of the dorsal ridge of a bird's bill (millimeters)          |
| Flipper Length    | an integer denoting the length penguin flipper (millimeters)                            |
| Body Mass         | an integer denoting the penguin body mass (grams)                                       |
| Sex               | a character string denoting the sex of an animal                                        |
| Delta 15 N        | a number denoting the measure of the ratio of stable isotopes 15N:14N                   |
| Delta 13 C        | a number denoting the measure of the ratio of stable isotopes 13C:12C                   |
| Comments          | a character string with text providing additional relevant information for data         |



## Create a String vector

### Excercise 1

Create a vector that contains at least three different characters

```{r create-vector, exercise=TRUE}

```

```{r create-vector-hint}
my_string <- c()
```

```{r create-vector-solution}
my_string <- c("my string", "has", "3 things!")
```

## stringr Functions

### Excercise 2

What function would you use to answer the following question:

*How many characters has the longest comment on the penguins_raw data set?*. 
(Remember the data set is called `penguins_raw` and the variable is called
`Comments`)

```{r number-characters, exercise=TRUE}

```

```{r number-characters-hint}
str_length()
```

```{r number-characters-solution}
str_length(penguins_raw$Comments)
```


### Exercise 3

We are going to clean the Species vector so we can have only the name of the species

First, we have to explore the `Species` vector. 

What function would you use to display on your console the information contained 
in the `Species` vector?

```{r explore, exercise=TRUE}


```

```{r explore-hint}
table(penguins_raw$Species)
#or
penguins_raw %>% 
  pull(Species) %>% 
  table()

```


### Excercise 4

Let's remove all the unnecessary information that is contained inside the `Species` vector
and create a new data set called penguins_clean


```{r clean-species,  exercise=TRUE}

```

```{r clean-species-hint}

penguins_clean <- 
  penguins_raw %>% 
  mutate(
    Species = str_remove_all(Species, "Penguin")
  ) %>% 
  mutate(
    Species = str_remove_all(Species,  "\\(.+")
   ) 

```

### Excercise 5

Your PI just realized that he does not like that the individual ID contains letters with 
that are not related to penguins. 
Modify the  letters of the code and replace it with a P in the data set you just created 
(`penguins_clean`)

```{r clean-id,  exercise=TRUE}

```

```{r clean-id-hint}
# penguins_clean <- 
#   penguins_raw %>% 
#   mutate(
#   `Individual ID` = str_replace_all( 
```

```{r clean-id-solution}
 penguins_clean <- 
  penguins_clean %>% 
  mutate(
  `Individual ID` = str_replace_all(
    `Individual ID`, 
    c("N" = "P", "A" = "P"))
  )
```


### Excercise 6

The vector named `Stage` has a comma separating the information. Explore the vector and separate 
it by the comma. The remaining information should not be eliminated, it should be stored in a new vector named `egg_stage`. To check your code select only the two variables of interest. 

```{r split, exercise=TRUE}

```


## Regular expressions

### Quiz

Some questions to verify that you understand the purposes of regular expressions

```{r quiz}
quiz(
  question("Which of the following symbols helps you escape in order to match a (?",
    answer("\\\\(", correct = TRUE),
    answer("\\//("),
    answer("("),
    answer("\\*(\\*")
  ),
  question("Which of the following symbols will match all the characters in a vector?",
    answer("D", correct = TRUE),
    answer("[az]"),
    answer("[AZ]"),
    answer("[A-Z]", correct = TRUE)
  )
)

```

