---
title: "Compare base and dplyr Subsetting"
author: "Gabriel Odom"
date: "10/2/2019"
output: 
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Overview
We will review

1. The data sets we have used so far in class
2. Row subsetting
3. Column subsetting
4. Relational subsetting

This lesson will not compare merging two data sets using basic R, because the dplyr version is superior in almost every way. Recall that we **join**ed part of the ACS data to the Miami-Dade ZIP Code data by the ZIP codes in both data sets.

*******************************************************************************

</br>

# Data Sets
In this class, we have used a few different data sets. The table below gives the names of the data sets and their sources.

| Data Set                    | Source                                        |
|-----------------------------|-----------------------------------------------|
| `mpg`                       | `tidyverse`                                   |
| `heroes_df`                 | In class: code below                          |
| `framingham_df`             | `LocalControl` with modifications shown below |
| `gapminder_df`              | `dslabs` with modifications shown below       |
| Miami-Dade County ZIP Codes | Blackboard: source in Lesson 08               |
| Broward County ZIP Codes    | Blackboard: source in Lesson 08               |
| 2016 ACS SNAP Data          | Blackboard: source in Lesson 08               |


## The `heroes_df` Data
```{r}
heroes_df <- tibble(
  subject_ID = factor(c("008", "016", "115", "027", "001")),
  name = c("Wonder Woman", "Green Lantern", "Spider-Man", "Batman", "Superman"),
  alias = c(
    "Diana Prince", "Alan Scott", "Peter Parker", "Bruce Wayne",
    "Clark Kent / Kal-El"
  ),
  city = c(
    "Gateway City", "Capitol City", "New York City", "Gotham", "Metropolis"
  ),
  male = c(FALSE, TRUE, TRUE, TRUE, TRUE),
  heightCM = c(183.5, 182.9, 177.8, 188.0, 190.5),
  weightKg = c(74.8, 91.2, 75.7, 95.3, 106.6),
  firstRun = c(1941L, 1940L, 1962L, 1939L, 1938L)
)
```

## Framingham Data
```{r}
library(LocalControl)

framingham_df <- 
  as_tibble(framingham) %>% 
  mutate(sex = ifelse(female == 1, yes = "Female", no = "Male")) %>% 
  mutate(curSmoke = cursmoke == 1) %>% 
  mutate(highBP = outcome == 1) %>% 
  mutate(death = outcome == 2) %>% 
  select(-female, -cursmoke, -outcome)
```

For the examples in this comparison document, we will use the first 100 records of the Framingham data only.
```{r}
framingham_df <- 
  framingham_df %>% 
  slice(1:100)
```


## Gapminder Data
```{r}
library(dslabs)

gapminder_df <-
  as_tibble(gapminder) %>% 
  mutate(gdpPerCap = gdp / population)
```

*******************************************************************************

</br>

# Row Subsetting

## base
To select the first row of a tibble using basic R subsetting rules, use the `[` function. This follows `row, column` syntax.
```{r}
framingham_df[1, ]
```

This returns a 1-row tibble.

## dplyr
To select the first row of a tibble using dplyr subsetting rules, use the `slice()` function. This uses row position only.
```{r}
framingham_df %>% 
  slice(1)
```

This returns a 1-row tibble.

*******************************************************************************

</br>

# Column Subsetting

## base
To select the first column of a tibble using basic R subsetting rules, again use the `[` function. This follows `row, column` syntax, so move the `1` on the other side of the comma.
```{r}
framingham_df[, 1]
```

This returns a 1-column tibble.

Because tibbles are also lists, we can use the list subsetting rules for columns as well. To subset by column position, we can use the `[` or the `[[` functions (for *single*-level or *double*-level subsetting). Note that `[[` can only take in a single scalar argument, rather than a vector of positions.
```{r}
framingham_df[1]
```

This returns a 1-column tibble.

```{r}
framingham_df[[1]]
```

This returns an atomic vector of the contents of the first column.

Note that these functions also work when we supply the name of a column as a character string:
```{r}
framingham_df[, "totchol"] # 1-column tibble
framingham_df["totchol"]   # atomic vector
framingham_df[["totchol"]] # 1-column tibble
```

One of the main differences between `[` and `[[` is that the single-subset operator can take in multiple column names, while `[[` cannot. For example:
```{r}
framingham_df[c("totchol", "death")]
```


Finally, we can use the `$` function to extract the contents of a *single column* by name *as a symbol* (not using quotes):
```{r}
framingham_df$totchol
```




## dplyr
To select the first column of a tibble using dplyr subsetting rules, use the `select()` function. This uses column position only.
```{r}
framingham_df %>% 
  select(1)
```

This returns a 1-column tibble.

To access a column by name, we have two options: `select()` and `pull()`. The `pull()` function works almost exactly like th `$` function: it takes in the name of a *single* column as a *symbol* object rather than a character string, and it returns the contents of the column as a vector.
```{r}
framingham_df %>% 
  pull(totchol)
```

We can also use the `select()` function. In some ways, this function works similarly to the `[` function: `select()` yields one or more columns of the tibble instead of the contents of a single column. However we do not supply the name in quotes, but rather as a symbol object. However, unlike the `$` function, .
```{r}
framingham_df %>% 
  select(totchol, death)
```

In order to extract the contents of a single column, use the `pull()` function instead:




*******************************************************************************

</br>

# Relational Subsetting
Finally, we can subset the rows of a data set by the relationships between the values a column and a specified value or by the relationships between one column and another.

## Example 1: Framingham Young Women
For example, we want to find all the females in the Framingham data set younger than the 75th percentile for the ages of these recorded females. (Remember that this lesson uses a subset of the Framingham data, not the full data set. Your results will look different.)

### Using base
```{r}
# Subset the women
framinghamWomen_df <- 
  framingham_df[framingham_df$sex == "Female", ]

# Find the age threshold for P75
womenP75_num <- quantile(framinghamWomen_df$age, 0.75)
framinghamWomenYoung_df <- 
  framinghamWomen_df[framinghamWomen_df$age < womenP75_num, ]

# Remove intermediary pieces from memory
rm(framinghamWomen_df, womenP75_num)

# Inspect
framinghamWomenYoung_df
```

### Using dplyr
```{r}
framingham_df %>%
  filter(sex == "Female") %>%
  filter(age < quantile(age, 0.75))
```


## Example 2: 
As another example, we want to find the 25 countries in the Gapminder data set which had the largest infant mortality rate in 2010.

### Using base
```{r}
# Save a subset of the Gapminder data only for the year 2011
gapminder2010_df <- 
  gapminder_df[gapminder_df[["year"]] == 2010, ]

# Find the order of the rows by worst infant mortality to best
worstInfMort_idx <- order(
  gapminder2010_df[["infant_mortality"]],
  decreasing = TRUE
)

# Subset the 2010 data by the first 25 indices in the "worst" vector
gapminderInfMort2010_df <-
  gapminder2010_df[worstInfMort_idx[1:25], ]

# Inspect
gapminderInfMort2010_df
```

### Using dplyr
```{r}
gapminder_df %>% 
  filter(year == 2010) %>% 
  arrange(desc(infant_mortality)) %>% 
  slice(1:25)
```

